working owls methods

### 2020-11-10 manuscript_working/ms_working_RMD.Rmd
updating methods with richness model

--- 
title: "working owls methods"
author: "Althea Archer"
date: "11/3/2020"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

For all three species' occupancy models, we ran 3 chains for 10,000 iterations and 1000 iterations discarded as burn-in, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. We visually inspected traceplots to verify that chains mixed well (Supplemental information S1-3xx).

>>>>

--- 
title: "working owls methods"
author: "Althea Archer"
date: "11/10/2020"
output:
  word_document:
    fig_caption: yes
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

For all three species' occupancy models, we ran 3 chains for 10,000 iterations and 1000 iterations discarded as burn-in, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. Initial values for $z_{t,h,i}$ were equivalent to one if an owl was detected in that survey $i$ and otherwise zero. We visually inspected traceplots to verify that chains mixed well (Supplemental information S1-3xx).

## Richness Model Framework

To model species richness at each route in each year, we assumed that there were 14 possible species present in El Salvador. This included the 10 species observed during our surveys (including unknown owl species) and an additional 4 unidentified species. This limit to species richness relates directly to the 13 species known to inhabit El Salvador (cite Owls of El Salvador here) plus one species we identified (Stygian owl), which was previously undocumented in El Salvador. We augmented our owl detection data with 4 additional potential species with all-zero detection [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,14$), $w_{t,h,s}$ is a binary indicator of that owl species being present in each route and year:

$$
w_{t,h,s} \sim \text{Bernoulli}(\Omega_h)
$$
where $\Omega_h$ is the probability of species presence in each route $h$. Now, the presence of an owl species in any route, year, and survey, $z_{t,h,i,s}$, is conditional on the probability of occupancy for that route, year, and species $\psi_{t,h,s}$ *and* that species belong to that route's community ($w_{t,h,s}$):

$$
z_{t,h,i,s} \sim \text{Bernoulli}(w_{t,h,s}*\psi_{t,h,s})
$$

Richness is conditional on each species presence and was estimated as the sum of owl species present in each route and year:

$$
\text{Richness}_{t,h} = \sum_{s=1}^{14} w_{t,h,s} 
$$

To share information about occupancy amongst species, we incorporated random effects into the parameterization of $\psi_{t,h,s}$:

\[
\begin{aligned}
\psi_{t,h,s} &\sim \text{Beta}(a^\psi_{t,h,s},b^\psi_{t,h,s})\\ 
\text{logit}(\mu^\psi_{t,h}) &= \text{E}(\psi_{t,h,s}) \sim N(0,2.25)\\
\ln(\rho^\psi) &\sim N(5,1)\text{, truncated to (0.01,10)}\\
a^\psi_{t,h,s} &= \mu^\psi_{t,h}\rho^\psi\\
b^\psi_{t,h,s} &= \rho^\psi - \mu^\psi_{t,h}\rho^\psi\\
\end{aligned}
\]

This parameterization allowed us to borrow strength across species in determining the mean probability of occupancy ($\mu^\psi_{t,h}$) for each route and year [@Guillera-Arroita:2019].

Consistent with our approach for the occupancy models, we assumed that detecting an owl depended on an owl of that species being present during that survey ($z_{t,h,i,s} = 1$) and the probability of detection $p_{t,h,i,j,k,s})$, which was related to the broadcast species:

$$
y_{t,h,i,j,k,s} \sim \text{Bernoulli}(z_{t,h,i,s}*p_{t,h,i,j,k,s})
$$

where generally $\text{logit}(p_{t,h,i,j,k,s}) = \beta_kX_k$ for each broadcast species $k$ as described above. The coefficients of the detection model were thus assumed to be consistent for all owl species, which allowed us to borrow information about the detection process for undetected owls from those that were detected.

### Richness Model Implementation

We used the R2jags package in R [@R:2014,@Plummer:2013] to implement the richness model. 
We ran 3 chains for 20,000 iterations and 2000 iterations discarded as burn-in and a thinning rate of 2, for a total of 27,000 iterations comprising the posterior distributions for each model parameter. Initial values for $z_{t,h,i,s}$ were equivalent to one if that species of owl $s$ was detected in that survey $i$ and otherwise zero. Similarly, initial values for $w_{t,h,s}$ were equivalent to one if that species of owl was detected in that route and year. We visually inspected traceplots to verify that chains mixed well (Supplemental information S4xx).
